---
title: "DMSD Vignette"
author: "Flávio Mota"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DMSD-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
*DMSD* is a package created to calculate Diversity Metrics from Spatialized Data. It is an alternative to process big datasets with less memory usage. The functions of this package can temporal beta diversity and its components turnover and nestedness for taxonomic, functional, and phylogenetic diversity. It also calculates alpha diversity to all three facets of diversity.

## Loading data
```{r, fig.height = 5, fig.width = 8, fig.align = 'center',}
library(terra)
library(BAT)
library(DMSD)

# Create SpatRaster objects
ref <- terra::rast(array(sample(c(rep(1, 800), rep(0, 200))), dim = c(10, 10, 10)))
names(ref) <- paste0("sp", 1:10)
ref

fut <- terra::rast(array(sample(c(rep(1, 400), rep(0, 600))), dim = c(10, 10, 10)))
names(fut) <- names(ref)
fut

# Create data.frame object
set.seed(100)
mass <- runif(10, 10, 800)
beak.size <- runif(10, .2, 5)
tail.length <- runif(10, 2, 10)
wing.length <- runif(10, 15, 60)
range.size <- runif(10, 10000, 100000)
traits <- data.frame(mass, beak.size, tail.length, wing.length, range.size)
rownames(traits) <- names(ref)
traits

# Create phylo object
set.seed(100)
tree <- ape::rtree(n = 10, tip.label = paste0("sp", 1:10))
tree
```

## Calculating alpha of taxonomic diversity
It is a spatial representation of *species richness* of a given scenario calculated by the sum of the same cell of each species.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# Alpha diversity for reference scenario
r.rich <- alpha(ref)
r.rich
terra::plot(r.rich, main = names(r.rich))
```

## Calculating beta temporal of taxonomic diversity
It is a spatial representation of *species composition* change between two scenarios. It indicates the differentiation in communities from the reference scenario to the future scenario by cell. Higher values of *beta diversity* means higher changes, which can be due to species replacement, *turnover*, or species gain/loss, *nestedness*.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# Beta total
beta.td <- beta.temporal(ref, fut)
beta.td
terra::plot(beta.td[[1]], main = names(beta.td[[1]]))

# Turnover
terra::plot(beta.td[[2]], main = names(beta.td[[2]]))

# Nestedness
terra::plot(beta.td[[3]], main = names(beta.td[[3]]))
```

## Calculating trait generalization
It is a spatial representation of the *average* of a given trait.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# Ecological generalization for the trait mass
mass.ref <- ecogen(ref, traits$mass) # reference scenario
mass.ref
terra::plot(mass.ref, main = names(mass.ref))

mass.fut <- ecogen(fut, traits$mass) # future scenario
mass.fut
terra::plot(mass.fut, main = names(mass.fut))
```
## Calculating alpha for functional diversity
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# Alpha functional diversity for reference scenario
r.rich.fd <- alpha(ref, traits)
r.rich.fd
terra::plot(r.rich.fd, main = names(r.rich.fd))
```

## Calculating beta temporal for functional diversity
It is a representation of *functional space* change between two scenarios. It indicates the differentiation in communities from the reference scenario to the future scenario by cell. Higher values of *functional beta diversity* means higher changes, which can be due to species replacement of functions, *turnover*, or  gain/loss of functions, *nestedness*.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# Beta total
beta.fd <- beta.temporal(ref, fut, traits)
beta.fd
terra::plot(beta.fd[[1]], main = names(beta.fd[[1]]))

# Turnover
terra::plot(beta.fd[[2]], main = names(beta.fd[[2]]))

# Nestedness
terra::plot(beta.fd[[3]], main = names(beta.fd[[3]]))
```
## Calculating alpha for phylogenetic diversity
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# Alpha functional diversity for reference scenario
r.rich.pd <- alpha(ref, tree)
r.rich.pd
terra::plot(r.rich.pd, main = names(r.rich.pd))
```

## Calculating beta temporal of phylogenetic diversity
It is a representation of the change in *phylogenetic relationship* of species between two scenarios. It indicates the differentiation in communities from the reference scenario to the future scenario by cell. Higher values of *phylogenetic beta diversity* means higher changes, which can be due to species replacement, *turnover*, or gain/loss, *nestedness*.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# Beta total
beta.pd <- beta.temporal(ref, fut, tree)
beta.pd
terra::plot(beta.pd[[1]], main = names(beta.pd[[1]]))

# Turnover
terra::plot(beta.pd[[2]], main = names(beta.pd[[2]]))

# Nestedness
terra::plot(beta.pd[[3]], main = names(beta.pd[[3]]))
```
