---
title: "DMSD Vignette"
author: "Flávio Mota"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DMSD-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
*DMSD* is a package created to calculate Diversity Metrics from Spatialized Data. It is an alternative to process big datasets with less memory usage. The functions of this package can calculate spatial and temporal beta diversity, separating the turnover and nestedness components of taxonomic, functional, and phylogenetic diversity.

## Loading data
```{r, fig.height = 5, fig.width = 8, fig.align = 'center',}
library(terra)
library(DMSD)

# SpatRaster
ref <- rast(system.file("extdata", "ref_Mota.tif", package = "DMSD"))
ref
fut <- rast(system.file("extdata", "fut_Mota.tif", package = "DMSD"))
fut

# data.frame
area <- read.csv(system.file("extdata", "area_Mota.csv", package = "DMSD"), sep = ";")
area$area2 <- area$area/1000 # adding new column
area$arealog <- log(area$area) # adding new column
area
```

## Calculating alpha diversity
It is a spatial representation of *species richness* of a given scenario calculated by the sum of the same cell of each species.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# alpha diversity for reference scenario
r.rich <- alphadv(ref)
r.rich
terra::plot(r.rich, main = names(r.rich))

# alpha diversity for future scenario
f.rich <- alphadv(fut)
f.rich
terra::plot(f.rich, main = names(f.rich))
```

## Calculating beta diversity
It is a spatial representation of *species composition* change between two scenarios. It indicates the differentiation in communities from the reference scenario to the future scenario by cell. Higher values of *beta diversity* means higher changes, which can be by species replacement, *turnover*, or species gain/loss, *nestedness*. The *ratio* is calculated dividing the *turnover* by the *total beta diversity*, thus, values close to 1 indicates predominance of *turnover* component while values close to 0 indicates predominance of *nestedness* component.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# total beta diversity between scenarios
beta <- betatempdv(ref, fut, "beta")
beta
terra::plot(beta, main = names(beta))

# turnover component
turn <- betatempdv(ref, fut, "turn")
turn
terra::plot(turn, main = names(turn))

# nestedness component
nest <- betatempdv(ref, fut, "nest")
nest
terra::plot(nest, main = names(nest))

# ratio (turnover/total beta diversity)
ratio <- betatempdv(ref, fut, "ratio")
ratio
terra::plot(ratio, main = names(ratio))

# calculate all beta diversity metrics at once
beta.all <- betatempdv(ref, fut)
beta.all
```

## Calculating trait generalization
It is a spatial representation of the *average* of a given trait.
```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# ecological generalization for the first trait
area1.t <- ecogen(ref, area, "one") # scenario 1
area1.t
terra::plot(area1.t, main = names(area1.t))

area2.t <- ecogen(fut, area, "one") # scenario 2
area2.t
terra::plot(area2.t, main = names(area2.t))

area.delta <- area2.t - area1.t # diference between scenarios
area.delta
terra::plot(area.delta)
```

```{r, fig.height = 4, fig.width = 5, fig.align = 'center',}
# ecological generalization for all column traits
traits <- ecogen(ref, area, "all")
traits
terra::plot(traits[[1]], main = names(traits[[1]]))
terra::plot(traits[[2]], main = names(traits[[2]]))
terra::plot(traits[[3]], main = names(traits[[3]]))
```
